Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-11-03 paper/graph.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@657 3fc257bb-8550-0410-81b3-a5c103da83f7

The single-source shortest path (SSSP) algorithm computes, for a
directed graph with weighted edges, the shortest weighted distance
from a chosen source vertex to all other vertices in the graph.  This
calculation is straightforward when global graph information is
available.  The source vertex is labeled with distance zero.  Then in
each iteration, edge adjacencies of labeled vertices are followed, and
adjacent vertices are relabeled with updated distances.  When no
distances change, the iterations are complete.  Most importantly, only
edges adjacent to modified vertices need be visited in each iteration.

For a distributed graph, global information about edge adjacencies and
labeled vertices is not available.  Instead, following the
Bellman-Ford-style algorithm of~\cite{SSSPMapReduce, Bellman58,
Ford62}, every edge of the graph is visited in each iteration,
although labels for an edge's vertices are updated only if the
breadth-first traversal from the source vertex has reached those
vertices.  A MapReduce formulation of this algorithm, described in
Figure~\ref{fig:sssp}, begins with a {\it map()} function loading
edges $(V_i, V_j)$ and source vertex $V_s$ with distance 0.  In each
iteration, the edges $(V_i, V_j)$ are collated with respect to $V_i$,
and a {\it reduce()} operation updates each $V_j$'s distance if
$V_i$'s distance is non-zero.

@@

while not $done$: \\
\> $done$ = 1 \\
\> Map $V$ using $E$ as input: add edge weights in $E$ to $V$ \\
\> \> input Key = $E_{ij}$, Value = $w_{ij}$ \\
\> \> output Key = $V_i$, Value = $[V_j, w_{ij}]$ \\
\> Collate $V$: Vertex is key \\
\> Reduce $V$:  Loop over edges of vertex to update distance to $V_s$ \\
\> \> Input: Key = $V_i$; Multivalue = edges $[V_j, w_{ij}]$ and candidate distances $d_i$ \\
\> \> Find smallest new distance $d_{min}$ for $V_i$ \\
\> \> If $d_{min} \ne d_i$: \\
\> \> \>  $done$ = 0 \\
\> \> \>  Output: Key = $V_j$, Value = $d_{min} + w_{ij}$ for each edge vertex $V_j$ \\
\> \>  Output:  Key = $V_i$, Value = $d_{min}$ \\
\> $done$ = MPI\_Allreduce($done$,MPI\_MIN)

@@

As with PageRank and R-MAT generation, significant savings in
communication and execution time can be achieved by pre-aggregating
certain MapReduce objects and storing data in more than one MapReduce
object.  The enhanced algorithm is shown in Figure~\ref{fig:sssp2}.
The MapReduce objects $V$ and $E$ storing vertices and edges are
aggregated across processors only once at the beginning of the
computation.  In each iteration of SSSP, only a small number of new
candidate distances are generated.  The enhanced algorithm uses
MapReduce object $U$ to store these updates.  Only the updates in $U$
are communicated; each update is aggregated to the same processor that
stores the corresponding vertex and its edges.  Thus, the total amount
of communication needed is smaller than in the original algorithm,
which aggregated all graph edges in each iteration.  Updated vertex
distances are stored in MapReduce object $V$, as before.

\begin{figure}[htb]
 \begin{center}\fbox{\begin{minipage}{\textwidth} \begin{tabbing}
  xxxx\=xxxx\=xxxx\=xxxx\=xxxx\=xxxxxxxxxxxxxxxxxxxxxxxxxx\= \kill

Pre-aggregated MapReduce object $V$ = graph vertices $V_i$ with distances $d_i$ from $V_s$:  \\
\> Key = $V_i$; Value = $\infty$.\\
Pre-aggregated MapReduce object $E$ containing graph edges ($V_i, V_j$) with weights $w_{ij}$: \\
\>  Key = $V_i$; Value = $[V_j, w_{ij}]$ \\
MapReduce object $U$ to store updates to distances = empty \\
Map $U$:  Add $V_s$ with distance $d_s = 0$ to $U$. \\
\> Input:  $V_s$ \\
\> Output:  Key = $V_s$; Value = 0. \\
done = 0 \\
while not $done$: \\
\> $done$ = 1 \\
\> Aggregate $U$ \\
\> Add $U$ to $V$, empty $U$ \\
\> Compress $V$:  Pick best candidate distance for each vertex \\
\> \> Input:  Key = $V_i$; Multivalue = candidate distances $d_i$ \\
\> \> Find smallest candidate distance $d_{min}$ for $V_i$ \\
\> \> \> If $d_{min} \ne d_i$, $done$ = 0 \\
\> \> Output to $V$:  Key = $V_i$; Value = $d_{min}$ \\
\> \> Output to $U$: if $d_{min} \ne d_i$, Key = $V_i$; Value = $d_{min}$ \\
\> $done$ = MPI\_Allreduce($done$,MPI\_MIN) \\
\> if not $done$: \\
\> \> Add $U$ to $E$.  Empty $U$. \\
\> \> Compress $E$:  Generate candidate distances for neighbors of changed vertices. \\
\> \> \> Input:  Key = $V_i$; Multivalue = $[V_j, w_{ij}]$ for edges ($V_i, V_j$) \\
\> \> \> \> and distance $d_i$ if updated in previous step. \\
\> \> \> Output to $U$:  If updated distance $d_i$ exists, \\
\> \> \> \> Key = $V_j$, Value = $d_i+w_{ij}$ for each edge ($V_i, V_j$) \\

@@

 \caption{\it Enhanced MapReduce algorithm for single-source shortest
 path (SSSP) calculation, with an additional MapReduce object $U$.
 Pre-aggregating the vertices $V$ and edges $E$, and later aggregating
 only updates $U$ to vertex distances, reduces communication and
 execution time.}

>>>>

For a directed graph with weighted edges, the single-source shortest
path (SSSP) algorithm computes the shortest weighted distance from a
chosen source vertex to all other vertices in the graph.  This
calculation is straightforward when global graph information is
available.  The source vertex is labeled with distance zero.  Then in
each iteration, edge adjacencies of labeled vertices are followed, and
adjacent vertices are relabeled with updated distances.  When no
distances change, the iterations are complete.  Most importantly, only
edges adjacent to modified vertices need be visited in each iteration.

For a distributed graph, global information about edge adjacencies and
labeled vertices is not available.  Instead, following the
Bellman-Ford-style algorithm of~\cite{SSSPMapReduce, Bellman58,
Ford62}, every edge of the graph is visited in each iteration and
vertex distancs are updated.  A MapReduce formulation of this
algorithm, described in Figure~\ref{fig:sssp}, stores directed edges
in $E$ and begins by labeling the distance to all vertices as
inifinte, except a distance of 0.0 to the source vertex $V_s$.  At
each iteration, a {\it map()} operation is performed using $E$ as
input to add edges and their weights to each vertex in $V$.  After a
{\it collate()} operation, each vertex $V_i$ knows its outgoing edges
and their weights, as well as its own current distance.  It may also
have one or more distances assigned to it from a previous iteration
(i.e. from an inbound edge).  In the {\it reduce()} operation, each
vertex resets its distance to the minimum of these new distances and
its current distance.  If its distance changed, it emits new distance
values $d_j$ for the vertices associated with its outgoing edges,
where $d_j = d_i + w_{ij}$.  It also unsets a ``done'' flag, stored by
each processor.  When the reduce is complete, an MPI\_Allreduce
operation is performed to check the ``done'' flag status on all
processors.  If any distance was changed, the iterations continue, and
the distance updates propagate through the graph.  If no distances
changed, the algorithm is done.

@@

while not $done$: \\
\> $done$ = 1 \\
\> Map $V$ using $E$ as input: add edge weights in $E$ to $V$ \\
\> \> input Key = $E_{ij}$, Value = $w_{ij}$ \\
\> \> output Key = $V_i$, Value = $[V_j, w_{ij}]$ \\
\> Collate $V$: Vertex is key \\
\> Reduce $V$:  Loop over edges of vertex to update distance to $V_s$ \\
\> \> Input: Key = $V_i$; Multivalue = edges $[V_j, w_{ij}]$ and candidate distances for $d_i$ \\
\> \> Find smallest new distance $d_{min}$ for $V_i$ \\
\> \> If $d_{min} \ne d_i$: \\
\> \> \>  $done$ = 0 \\
\> \> \>  Output: Key = $V_j$, Value = $d_{min} + w_{ij}$ for each edge vertex $V_j$ \\
\> \>  Output:  Key = $V_i$, Value = $d_{min}$ \\
\> $done$ = MPI\_Allreduce($done$,MPI\_MIN)

@@

As with PageRank and R-MAT generation, significant savings in
communication and execution time can be achieved by pre-aggregating
the data stored in MapReduce objects and by splitting the data
appropriately across more than one MapReduce object.  An enhanced SSSP
algorithm is shown in Figure~\ref{fig:sssp2}.  The MapReduce objects
$V$ and $E$ which store vertices and edges are aggregated across
processors only once at the beginning of the computation, which means
the communication portion of the {\it collate()} operation is
performed to assign each key to a processor.  The {\it collate()}
operation of Figure~\ref{fig:sssp} can then be replaced with a {\it
compress()} operation which requires no communication.

In each iteration of SSSP, typically only a small number of new
distances are generated.  The enhanced algorithm uses MapReduce object
$U$ to store these updates.  Only the updates in $U$ are communicated;
which aggregates them to the same processor that permanently stores
the corresponding vertex and its edges.  Thus, the total amount of
required communication is smaller than in the original algorithm,
which communicated all graph edges at each iteration.  Updated vertex
distances are stored in MapReduce object $V$, as before.

\begin{figure}[htb]
 \begin{center}\fbox{\begin{minipage}{\textwidth} \begin{tabbing}
  xxxx\=xxxx\=xxxx\=xxxx\=xxxx\=xxxx\=xxxxxxxxxxxxxxxxxxxxxx\= \kill

Pre-aggregated MapReduce object $V$ = graph vertices $V_i$ with distances from sourece vertex $V_s$:  \\
\> Key = $V_i$; Value = $\infty$, except Value for $V_s$ = 0 \\
Pre-aggregated MapReduce object $E$ = graph edges with weights: Key = ($V_i, V_j$), Value = $w_{ij}$ \\
MapReduce object $U$ to store updates to distances, initially empty \\
done = 0 \\

while not $done$: \\
\> $done$ = 1 \\
\> Aggregate $U$ \\
\> Add $U$ to $V$, empty $U$ \\
\> Compress $V$:  Compute best distance for each vertex \\
\> \> Input:  Key = $V_i$; Multivalue = candidate distances for $d_i$ \\
\> \> Find smallest new distance $d_{min}$ for $V_i$ \\
\> \> If $d_{min} \ne d_i$: \\
\> \> \> $done$ = 0 \\
\> \> \> Output to $U$: Key = $V_i$; Value = $d_{min}$ \\
\> \> Output to $V$: Key = $V_i$; Value = $d_{min}$ \\
\> $done$ = MPI\_Allreduce($done$,MPI\_MIN) \\
\> if not $done$: \\
\> \> Add $U$ to $E$, empty $U$ \\
\> \> Compress $E$: Generate candidate distances for neighbors of changed vertices. \\
\> \> \> Input:  Key = $V_i$; Multivalue = $[V_j, w_{ij}]$ and candidate distance $d_i$ if updated in previous step \\
\> \> \> If updated distance $d_i$ exists: \\
\> \> \> \> Output to $U$: Key = $V_j$, Value = $d_i+w_{ij}$ for each edge vertex $V_j$) \\

@@

 \caption{\it Enhanced MapReduce algorithm for single-source shortest
 path (SSSP) calculation, with an additional MapReduce object $U$.
 Pre-aggregating the vertices $V$ and edges $E$, and storing only
 changed distances in $U$, reduces communication and execution time.}

### 2010-11-03 paper/lessons.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@655 3fc257bb-8550-0410-81b3-a5c103da83f7
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@657 3fc257bb-8550-0410-81b3-a5c103da83f7

MapReduce achieves parallelism through randomizing the distribution of
data across processors, which often intentionally ignores data
locality.  This translates into maximal data movement (during a data
shuffle or {\it collate()} operation) with communication between all
pairs of processors.  But the benefit is often good load-balance, even
for hard-to-balance irregular data sets.  By contrast, more
traditional distributed-memory parallel algorithms, e.g. for matrix
operations, or grid-based or particle-based simulation codes,
typically work hard to localize data and minimize communication.  To
do this they often require a lot of application-specific logic and
parallel communication coding to create and maintain a data
decomposition, generate ghost versions of nearby spatially-decomposed
data, etc.

MapReduce algorithms can be hard to design, but are often relatively
easy to write and debug.  Thinking about a computational task from a
MapReduce perspective is different from traditional distributed-memory
parallel algorithm design.  For example, with an MPI mindset, it often
seems heretical to intentionally ignore data locality.  However,
writing small {\it map()} and {\it reduce()} functions is typically
easy.  And writing an algorithm that involves complex parallel
operations without needing to write application-specific code to
communicate data via MPI calls or to move data between out-of-core
disk files and processor memory is often a pleasant surprise.
Moreover, if the MapReduce algorithm is initially coded so that it
runs correctly on one processor, it often works out-of-the-box on
hundreds or thousands of processors, without the need for additional
debugging.

Performing MapReduce operations on a fixed allocation of processors on
a traditional MPI-based parallel machine is a somewhat different
conceptual model than that of performing MapReduce operations on a
cloud computer using (for example) Hadoop.  In the former case, one
can potentially control which processor owns which data at various
stages of an algorithm.  This is somewhat hidden from the user in a
typical cloud-computing model, where data simply exists somewhere in
the cloud and Hadoop ensures data moves where it is needed and is
operated on by some processor.  The cloud model is a nice data-centric
abstraction which allows for fault tolerance both to data loss (via
redundant storage) and to processor failure (via reassignment of
work), neither of which is typically possible on current MPI-based
parallel machines.

>>>>

MapReduce achieves parallelism through randomizing the distribution of
data across processors, which often intentionally ignores data
locality.  This translates into maximal data movement (during a data
shuffle or {\it collate()} operation) with communication between all
pairs of processors.  But the benefit is often good load-balance, even
for hard-to-balance irregular data sets, such as those derived from
sparse graphs with power-law degree distributions.  By contrast, more
traditional distributed-memory parallel algorithms, e.g. for matrix
operations, or grid-based or particle-based simulation codes,
typically work hard to localize data and minimize communication.  To
do this they often require application-specific logic and parallel
communication coding to create and maintain a data decomposition,
generate ghost copies of nearby spatially-decomposed data, etc.

MapReduce algorithms can be hard to design, but are often relatively
easy to write and debug.  Thinking about a computational task from a
MapReduce perspective is different from traditional distributed-memory
parallel algorithm design.  For example, with an MPI mindset, it often
seems heretical to intentionally ignore data locality.  However,
writing small {\it map()} and {\it reduce()} functions is typically
easy.  And writing an algorithm that involves complex parallel
operations without needing to write application-specific code to
communicate data via MPI calls or to move data between out-of-core
disk files and processor memory is often a pleasant surprise.  By
contrast, extending a library like Trilinos (or other in-core MPI
applications) to enable it to perform its suite of linear algebra
operations in out-of-core mode on large data sets would be a
considerable effort.  Moreover, if a MapReduce algorithm is initially
coded so that it runs correctly on one processor, it often works
out-of-the-box on hundreds or thousands of processors, without the
need for additional debugging.

Performing MapReduce operations on a fixed allocation of processors on
a traditional MPI-based parallel machine is a somewhat different
conceptual model than performing MapReduce operations on a cloud
computer using (for example) Hadoop.  In the former case, one can
potentially control which processor owns which data at various stages
of an algorithm.  This is somewhat hidden from the user in a typical
cloud-computing model, where data simply exists somewhere in the cloud
and Hadoop ensures data moves where it is needed and is operated on by
some processor.  The cloud model is a nice data-centric abstraction
which allows for fault tolerance both to data loss (via redundant
storage) and to processor failure (via reassignment of work), neither
of which is typically possible on current MPI-based parallel machines.
